---
title: "Time Series Shark Attacks"
author: "Josh Balson"
date: "7/28/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Sets

<https://www.sharks.org/global-shark-attack-file>        
Filtered for USA only       
Filtered to include the time frame Jan 1972 - Dec 2020        

```{r init, include=FALSE}
rm(list = ls())
library(dplyr)
library(tswge)
library(tidyr)
library(forecast)
library(tseries)
library(xts)
library(orcutt)
library(nnfor)
library(vars)
library(plotrix)
path = 'C:\\Users\\balso\\OneDrive\\Documents\\Summer2021\\TimeSeries\\Sharks\\'
```

```{r udfs, include=FALSE}
source(paste(path,'udf_make_icecream.R', sep=''))
source(paste(path,'udf_make_data.R', sep=''))
source(paste(path,'udf_make_models.R', sep=''))
source(paste(path,'udf_make_all.R', sep=''))
source(paste(path,'udf_make_plots.R', sep=''))
source(paste(path,'udf_get_all_ase.R', sep=''))
```

```{r get sharks}
sharks <- read.csv("sharks.csv", header = TRUE)
head(sharks)
```


```{r get ice cream}
ice.cream <- udf_make_icecream('IceCreamIndex.csv')
head(ice.cream)
```

## Data Preprocessing
```{r}
data <- udf_make_data(sharks, ice.cream)

#ts1
ts.sharks <- data[[1]]
ts.sharks.train <- data[[2]]
ts.sharks.test <- data[[3]]

#ts2
ts.ice.cream <- data[[4]]
ts.ice.cream.train <- data[[5]]
ts.ice.cream.test <- data[[6]]
```


## Plot data
```{r plot1}
plot(as.xts(ts.sharks), major.format='%Y-%m',main='Shark Attacks United States',ylim=c(0,25))
```

ACF and Spectral Density
```{r plot2, results=FALSE}
plotts.sample.wge(ts.sharks.train)
```

## Build Forecast Models
```{r message=FALSE}
models <- udf_make_models(ts.sharks.train, ts.sharks.test, ts.ice.cream, ts.ice.cream.train)
```

Seasonal Model    
ARIMA(15,3) with s = 12  
$$
(1-0.3645 B+ 0.3245 B^{ 2 }+ 0.5476 B^{ 3 }-0.0081 B^{ 4 }-0.0304 B^{ 5 }+ 0.0216 B^{ 6 }...
$$
$$
...-0.06 B^{ 7 }-0.0301 B^{ 8 }-0.0137 B^{ 9 }-0.0364 B^{ 10 }-0.0781 B^{ 11 }+ 0.4743 B^{ 12 }-0.318 B^{ 13 }+ 0.1924 B^{ 14 }+ 0.1346 B^{ 15 })X_t = ...
$$
$$
...(1-0.2669B-0.2659238B^{ 2 } +0.7336743B^{ 3 })a_t
$$

$$
\sigma^2 = 6.0746
$$

Plot Seasonal 24 Month Forecast
```{r}
plot(ts.sharks[541:588],main='Shark Attacks Seasonal Forecast - 24 Months'
     ,ylim=c(0,25), ylab='Shark Attacks',type='o', xlab='Months')
points(models[[3]][541:588],type='o',lty=1,lwd=3,col=2,cex=.6)
```

Get 24 Month ASE for Seasonal     
```{r f_seas24 ASE}
ASE_seas <- models[[2]]
ASE_seas
```

Get 12 Month ASE for Seasonal     
```{r f_seas12 ASE}
ASE_seas12 <- models[[4]]
ASE_seas12
```

Stationary Model
ARMA(3,3)
$$
(1-2.7255B+2.7211B^{ 2 }-0.9942B^{ 3 })(X_t - 2.828)=(1-2.6349307B+2.5456529B^{ 2 }-0.8923811B^{ 3 })a_t
$$

Plot Stationary 24 Month Forecast
```{r}
plot(ts.sharks[541:588],main='Shark Attacks Stationary Forecast - 24 Months'
     ,ylim=c(0,25), ylab='Shark Attacks',type='o', xlab='Months')
points(models[[7]][541:588],type='o',lty=1,lwd=3,col=2,cex=.6)
```

Get 24 Month ASE for Stationary     
```{r f_stat24 ASE}
ASE_stat <- models[[6]]
ASE_stat
```

Get 12 Month ASE for Stationary     
```{r f_stat12 ASE}
ASE_stat12 <- models[[8]]
ASE_stat12
```

Signal Plus Noise
```{r}
# test signal significance
summary(models[[9]])
```

```{r}
# cochrane-orcutt test
summary(models[[10]])
```

Signal Plus Noise Model
Linear with ARMA(7,4)
$$
X_t = s_t + Z_t ,where
$$
$$
s_t = 0.0093272*t+0.1930892, and 
$$

$$
Z_t = (1-0.2826B^{1}-0.6936B^{2}-0.1279B^{3}+1.0027B^{4}-0.073B^{5}+0.0328B^{6}...     
$$

$$
...-0.0748B^{7})(X_t-2.828014)-(1-0.139B^{1}-0.7524B^{2}-0.1588B^{3}+0.9437B^{4})a_t     
$$

```{r}
plot(ts.sharks[541:588],main='Shark Attacks SPN Forecast - 24 Months'
     ,ylim=c(0,25), ylab='Shark Attacks',type='o', xlab='Months')
points(models[[14]][541:588],type='o',lty=1,lwd=3,col=2,cex=.6)
points(models[[12]][541:588],type='l',lty=3,lwd=2,col=4,cex=.6)
```

Get 24 Month ASE for SPN     
```{r f_spn24 ASE}
ASE_spn <- models[[13]]
ASE_spn
```

Get 12 Month ASE for SPN     
```{r f_spn12 ASE}
ASE_spn12 <- models[[15]]
ASE_spn12
```

## Bivariate Models
VAR   
MLP     
Neural Network    

View New Data - Ice Cream Production Index    
<https://fred.stlouisfed.org/series/IPN31152N>    

```{r ice cream plot1}
plot(as.xts(ts.ice.cream), major.format='%Y-%m',main='Ice Cream Production Index United States')
```

Look at ACF and Spectral Density for Ice Cream Production
```{r, results=FALSE}
plotts.sample.wge(ts.ice.cream.train)
```

View Both Data Sets
```{r}
par(mfrow=c(2,1))
plot(as.xts(ts.sharks), main='Shark Attacks, U.S')
plot(as.xts(ts.ice.cream), main='Ice Cream Production Index, U.S')
```

Zoom-in Past Seven Years - Log Transforms
```{r}
par(mfrow=c(2,1))
plot(as.xts(log1p(ts.sharks))[503:588], main='Log Shark Attacks, U.S')
plot(as.xts(log1p(ts.ice.cream))[503:588], main='Log Ice Cream Production Index, U.S')
```

VAR model with Ice Cream Production
```{r}
plot(ts.sharks[541:588],main='Shark Attacks VAR Forecast - 24 Months'
     ,ylim=c(0,25), ylab='Shark Attacks',type='o', xlab='Months')
points(models[[18]][541:588],type='o',lty=1,lwd=3,col=2,cex=.6)
```

```{r}
fanchart(models[[16]])
```

Get 24 Month ASE for VAR     
```{r f_var24 ASE}
ASE_VAR <- models[[17]]
ASE_VAR
```

Get 12 Month ASE for SPN     
```{r f_var12 ASE}
ASE_VAR12 <- models[[19]]
ASE_VAR12
```

MLP model with Ice Cream Production
```{r}
models[[20]]
plot(models[[20]])
```

```{r}
plot(ts.sharks[541:588],main='Shark Attacks MLP Forecast - 24 Months'
     ,ylim=c(0,25), ylab='Shark Attacks',type='o', xlab='Months')
points(models[[22]][541:588],type='o',lty=1,lwd=3,col=2,cex=.6)
```

Get 24 Month ASE for MLP     
```{r f_mlp24 ASE}
ASE_MLP <- models[[21]]
ASE_MLP
```

Get 12 Month ASE for MLP     
```{r f_mlp12 ASE}
ASE_MLP12 <- models[[23]]
ASE_MLP12
```

```{r}
plot(ts.sharks,main='Shark Attacks MLP Forecast - 24 Months'
     ,ylim=c(0,25), ylab='Shark Attacks',type='o', xlab='Months')
points(models[[22]],type='o',lty=1,lwd=3,col=2,cex=.6)
```

## Generate Forecasts Using Different Time Ranges for Training

(note: plots from artrans.wge could not be suppressed using 'plottr = F' when called by the user defined functions)
```{r message=FALSE}
forecast_years <- list(5,10,20,30,40,47)
# forecast_years <- list(5,10)
ln <- F
modelsAll <- list()

for (f in forecast_years){
  
  data <- udf_make_data(sharks, ice.cream, f, 2, ln = ln)
  
  m <- udf_make_models(data[[2]], data[[3]], data[[4]], data[[5]], ln = ln)
  
  tmp <- m
  
  modelsAll <- append(modelsAll,list(tmp))
}
```

## Generate Log Forecasts Using Different Time Ranges for Training

(note: plots from artrans.wge could not be suppressed using 'plottr = F' when called by the user defined functions)
```{r message=FALSE}
ln <- T
modelsAllLog <- list()

for (f in forecast_years){
  
  data <- udf_make_data(sharks, ice.cream, f, 2, ln = ln)
  
  m <- udf_make_models(data[[2]], data[[3]], data[[4]], data[[5]], ln = ln)
  
  tmp <- m
  
  modelsAllLog <- append(modelsAllLog,list(tmp))
}
```

## Compile ASEs for all Models

ASEs in ascending order (12 month)
```{r, echo=FALSE}
all_ase <- udf_get_all_ase()
print(all_ase[order(all_ase$ASE12),])
# write.csv(all_ase,paste(path,'ts_model_ase.csv', sep=''), row.names=F)
```


Look at Top Six Models Using 12 Month ASE
```{r}
par(mfrow=c(2,3))
udf_make_plots(sharks, modelsAll[[5]][[22]], yrs_train = 40, title = '1.) MLP 40 Years')
udf_make_plots(sharks, modelsAll[[6]][[22]], yrs_train = 47, title = '2.) MLP 47 Years')
udf_make_plots(sharks, modelsAll[[4]][[22]], yrs_train = 30, title = '3.) MLP 30 Years')
udf_make_plots(sharks, modelsAllLog[[3]][[22]], yrs_train = 20, title = '4.) Log MLP 20 Years')
udf_make_plots(sharks, modelsAll[[5]][[18]], yrs_train = 40, title = '5.) VAR 40 Years')
udf_make_plots(sharks, modelsAll[[2]][[14]], yrs_train = 10, title = '6.) SPN 10 Years')
```

Look at All Seasonal Models by 12 Month ASE
```{r}
par(mfrow=c(2,3))
udf_make_plots(sharks, modelsAll[[4]][[3]], yrs_train = 30, title = '1.) Seasonal 30 Years')
udf_make_plots(sharks, modelsAll[[5]][[3]], yrs_train = 40, title = '2.) Seasonal 40 Years')
udf_make_plots(sharks, modelsAll[[6]][[3]], yrs_train = 47, title = '3.) Seasonal 47 Years')
udf_make_plots(sharks, modelsAll[[2]][[3]], yrs_train = 10, title = '4.) Seasonal 10 Years')
udf_make_plots(sharks, modelsAll[[1]][[3]], yrs_train = 5, title = '5.) Seasonal 5 Years')
udf_make_plots(sharks, modelsAll[[3]][[3]], yrs_train = 20, title = '6.) Seasonal 20 Years')
```

Look at All Seasonal Log Models by 12 Month ASE
```{r}
par(mfrow=c(2,3))
udf_make_plots(sharks, modelsAllLog[[4]][[3]], yrs_train = 30, title = '1.) Log Seasonal 30 Years')
udf_make_plots(sharks, modelsAllLog[[5]][[3]], yrs_train = 40, title = '2.) Log Seasonal 40 Years')
udf_make_plots(sharks, modelsAllLog[[6]][[3]], yrs_train = 47, title = '3.) Log Seasonal 47 Years')
udf_make_plots(sharks, modelsAllLog[[2]][[3]], yrs_train = 10, title = '4.) Log Seasonal 10 Years')
udf_make_plots(sharks, modelsAllLog[[1]][[3]], yrs_train = 5, title = '5.) Log Seasonal 5 Years')
udf_make_plots(sharks, modelsAllLog[[3]][[3]], yrs_train = 20, title = '6.) Log Seasonal 20 Years')
```

Look at Top Six Stationary Models by 12 Month ASE
```{r}
par(mfrow=c(2,3))
udf_make_plots(sharks, modelsAll[[5]][[7]], yrs_train = 40, title = '1.) Stationary 40 Years')
udf_make_plots(sharks, modelsAll[[4]][[7]], yrs_train = 30, title = '2.) Stationary 30 Years')
udf_make_plots(sharks, modelsAll[[6]][[7]], yrs_train = 47, title = '3.) Stationary 47 Years')
udf_make_plots(sharks, modelsAll[[2]][[7]], yrs_train = 10, title = '4.) Stationary 10 Years')
udf_make_plots(sharks, modelsAllLog[[3]][[7]], yrs_train = 20, title = '5.) Log Stationary 20 Years')
udf_make_plots(sharks, modelsAll[[1]][[7]], yrs_train = 5, title = '6.) Stationary 5 Years')
```
## Ensemble Model

Look at one ensemble model
Use lower amplitude stationary forecasts to ensemble with Log MLP forecast
```{r}
fc_mlpLog20yrs <- modelsAllLog[[3]][[22]][241:264]
fc_stat10yrs <- modelsAll[[2]][[7]][121:144]

ensemble <- (fc_mlpLog20yrs + (.5 * fc_stat10yrs)) / 2 # wtd avg. stationary 1/2 the weight 

t <- 1:length(ensemble)

plot(t,fc_mlpLog20yrs,main='Shark Attacks Ensemble Forecast - MLP and Stationary'
     ,ylim=c(0,25), ylab='Shark Attacks',type='l',lty=2, xlab='Months', col=4,lwd=2)
points(t,ensemble,type='o',lty=1,lwd=3,col=2,cex=.6)
points(t,0.5*fc_stat10yrs,type='l',lty=2, col=6,lwd=2)
legend(2,25, legend=c('LogMLP','Ensemble','Stationary'), col=c(4,2,6), lty=2:1)
```

## Final Model Forecast - Piecewise

Ensemble for months 1 - 12    
Log MLP 20 Years for months 13 - 24   

where Ensemble is the weighted average of Log MLP 20 year and Stationary 10 year

$$
forecast_p = modelA + modelB    
$$
$$
where
$$

$$
modelA= MLPLog20yrs(1:12)+(1/2)Stationary10yrs(1:12)
$$
$$
and
$$

$$
modelB= MLPLog20yrs(13:24)
$$
Plot Final Forecast
```{r}
piecewise <- c(ensemble[1:12],fc_mlpLog20yrs[13:24])
# 24 month

# expand time for plotting
piecewise <- c(rep(NA,288),piecewise)
piecewise <- piecewise[(length(piecewise) - 108 + 1): length(piecewise)]

# get actual data, line up for plotting
ts.sharks.20yr <- ts.sharks[(length(ts.sharks) - 84 + 1): length(ts.sharks)]
ts.sharks.20yr <- c(ts.sharks.20yr,rep(NA,24))

# plot final model with CIs
t <- 1:length(ts.sharks.20yr)

# get CIs
CU <- piecewise + 1.96 * std.error(piecewise)
CU <- c(CU[1:83],0,CU[85:108])
CL <- piecewise - 1.96 * std.error(piecewise)
CL[CL<0] <- 0
CL <- c(CL[1:83],0,CL[85:108])

plot(t,ts.sharks.20yr,main='Shark Attacks Final 24 Month MLP Forecast'
     ,ylim=c(0,25), ylab='Shark Attacks',type='o', xlab='Months')
points(t,CL,type='l',lty=2)
points(t,piecewise,type='o',lty=1,lwd=3,col=2,cex=.6)
points(t,CU,type='l',lty=2)
legend(2,25, legend=c('Upper Limit','Forecast','Lower Limit'), col=c(1,2,1), lty=2:1)
```
Total Shark Attacks Projected in the U.S - 2021
```{r}
f_2021 <- round(ts(piecewise[!is.na(piecewise)][1:12], frequency = 12, start = 2021),0)
print(f_2021)
print(paste('2021 Total Shark Attacks Projected -',sum(f_2021), sep=' '))
```

Total Shark Attacks Projected in the U.S - 2022
```{r}
f_2022 <- round(ts(piecewise[!is.na(piecewise)][13:24], frequency = 12, start = 2022),0)
print(f_2022)
print(paste('2022 Total Shark Attacks Projected -',sum(f_2022), sep=' '))
```
Bring Prior Three Years of Actual Shark Attacks for Forecast Perspective
```{r}
act_2020 <- sum(ts.sharks.20yr[!is.na(ts.sharks.20yr)][73:84])
act_2019 <- sum(ts.sharks.20yr[!is.na(ts.sharks.20yr)][61:72])
act_2018 <- sum(ts.sharks.20yr[!is.na(ts.sharks.20yr)][49:60])

actualPast3 <- ts(c(act_2018,act_2019,act_2020),frequency=1,start=2018)
proj2 <- ts(c(sum(f_2021),sum(f_2022)),frequency=1,start=2021)

print(actualPast3)
print(proj2)
```

Upper Limits - Total Shark Attacks Projected in the U.S - 2021
```{r}
f_2021 <- round(ts(CU[!is.na(CU)][1:12], frequency = 12, start = 2021),0)
print(f_2021)
print(paste('2021 Total Shark Attacks Projected -',sum(f_2021), sep=' '))
```
Lower Limits - Total Shark Attacks Projected in the U.S - 2021
```{r}
f_2021 <- round(ts(CL[!is.na(CL)][1:12], frequency = 12, start = 2021),0)
print(f_2021)
print(paste('2021 Total Shark Attacks Projected -',sum(f_2021), sep=' '))
```

Upper Limits - Total Shark Attacks Projected in the U.S - 2022
```{r}
f_2022 <- round(ts(CU[!is.na(CU)][13:24], frequency = 12, start = 2022),0)
print(f_2022)
print(paste('2022 Total Shark Attacks Projected -',sum(f_2022), sep=' '))
```

Lower Limits - Total Shark Attacks Projected in the U.S - 2022
```{r}
f_2022 <- round(ts(CL[!is.na(CL)][13:24], frequency = 12, start = 2022),0)
print(f_2022)
print(paste('2022 Total Shark Attacks Projected -',sum(f_2022), sep=' '))
```





